Fast gravitational wave parameter estimation without compromises

### 2023-07-21 src/tex/ms.tex
Change software citation to \software

% Old codes have trouble handling next gen data
% The PE results in the most recent catalog of GW events are produced by several
% community-developed PE codes
% \mi{what catalog do you mean? The LVK doesn't use PyCBC Inference}
There exist a number of prominent, community-developed PE codes, including
\textsc{LALInference} \citep{Veitch:2014wba}, \textsc{PyCBC Inference}
\citep{Biwer:2018osg}, and \textsc{Bilby}
\citep{Ashton:2018jfp,Romero-Shaw:2020owr}.  These packages have been tested by a
number of groups and are well regarded as standard tools. However, while these
tools have passed many robustness tests, they are known to be computationally
intensive. The exact amount of time needed to analyze one event depends on
factors like the duration and frequency of the signal, as well as features of
the specific waveform model. Typical runtimes for production-level analyses can
range from hours to weeks.

@@

To obtain a reference waveform, we currently use the \textsc{differential
evolution} algorithm \citep{Storn1997DifferentialE} available in the
\textsc{scipy} package \citep{2020SciPy-NMeth} to find the waveform parameters
which maximize the likelihood.  The reference waveform could also be produced
from trigger parameters precomputed by a search pipeline without additional
computation. Once we have obtained a reference waveform, we can check the accuracy
of the heterodyned likelihood by comparing its value to the original likelihood
at several points in the parameter space. We can then choose the number of bins such
that the difference between the values of the two likelihoods is smaller than
chosen tolerance threshold.

@@

While transition kernels that use gradient information can help improve the
acceptance rate, computing the gradient of the posterior density function
introduces an additional computational cost, which is not necessarily
beneficial in terms of sampling time. If one wants to compute the gradient
through finite differencing, the additional computational cost goes as at least
${\sim} \mathcal{O}(2n)$, where $n$ is the dimension of the problem. On the
other hand, schemes like \textsc{Jax}
\citep{jax2018github} allow us to compute the gradient of the likelihood
function with respect to the parameters through automatic differentiation,
which gives the gradient information down to machine precision at around the
same order of time compared to evaluating the posterior itself. Thus, having
access to gradient information through automatic differentiation is crucial to
making gradient-based transition kernels favorable in terms of computing cost.

To leverage automatic differentiation, the entire posterior function must be
implemented within \textsc{Jax} or a similar framework; this includes the
likelihood and, therefore, the waveform approximant.  This means that the
development of differentiable approximants is essential for leveraging
gradient-based sampling in GW applications. We currently make use of the
waveforms implemented in \textsc{ripple} \citep{ripplepaper}.

@@

Note that every time we retrain the network, we are breaking the Markov
properties since we are changing the proposal distribution. To produce final
samples that can be used to estimate target quantities, one has to freeze the
normalizing flow model and not retrain during the final sampling phase in order
to satisfy the detailed balance condition. We use the package \textsc{flowMC}
\citep{2022arXiv221106397W,Gabrie:2021tlu}, with MALA as the gradient-based sampler. The
pseudocode of the algorithm is given in Algorithm \ref{alg:cap}.

\begin{algorithm}
\caption{\textsc{flowMC} pseudocode}\label{alg:cap}
\KwIn{initial position $ip$}
\Parameters{number of training loops $nt$, number of production loops $np$}
\Variables{current chains $cc$, current position $cp$, current NF parameters $\Theta$, chains from local sampler $c_{local}$, chains from global sampler $c_{global}$}
\KwResult{output chains $chains$}
$cp \leftarrow ip$\\
\tcc{Training loop}
    \For{$i<nt$}{
        $cc, cp \leftarrow LocalSampling(cp)$\\
        $\Theta \leftarrow TuneNF(cc)$\\
        $c_{global}, cp \leftarrow GlobalSampling(cp, \Theta)$ \\
        $cc \leftarrow Append(cc, c_{global})$
    }
\tcc{Production loop}
    \For{$i<np$}{
        $c_{local}, cp \leftarrow LocalSampling(cp)$\\
        $c_{global}, cp \leftarrow GlobalSampling(cp, \Theta)$ \\
        $chains \leftarrow Append(chains, c_{local}, c_{global})$
    }

@@

% GPU helps packing a shit ton of waveform evaluation
Another benefit of accelerators is the parallel evaluation of waveforms. Since
the waveform model we use can be evaluated at any given time or frequency
independently, this means computing a waveform can be trivially parallelized
over frequency bins. On an \textsc{Nvidia} A100 GPU, we can evaluate the
waveform model $\sim \mathcal{O}(10^9)$ times in a second for different
frequencies or source parameters. Together with the heterodyned likelihood, this
allows us to run thousands of parallel chains in a PE run. The high throughput
of waveform evaluations unlocks the potential of the \textsc{flowMC} sampling
algorithm.

@@

% State the distribution of injected population
We sample 1200 events from the distribution of parameters detailed in Table
\ref{tab:parameters}; the same distributions are used as the prior in the PE
process.  We simulate signals over $16\,$s of data, with a minimum frequency
cutoff of $30\,$Hz and a sampling rate of $2048\,$Hz. We draw noise from a set of
projected design PSDs for the LIGO Hanford, LIGO Livingston
(\texttt{SimNoisePSDaLIGOZeroDetHighPower}) and Virgo
(\texttt{SimNoisePSDAdvVirgo}) detectors
\citep{lalsuite,Shoemaker:T0900288,2012arXiv1202.4031M}. For both injection and
recovery, we make use of the \textsc{IMRPhenomD} waveform \citep{Khan:2015jqa}
via the fully-differentiable implementation presented in the \textsc{ripple}
package \citep{ripplepaper}. We use a neural spline flow model \citep{2019arXiv190604032D}
with 10 layers, each with 128 hidden units, and 8 bins per layer as our
normalizing flow model.

@@

To demonstrate the performance of our parameter estimation pipeline, we apply
it to two real LIGO-Virgo events: GW150914 and GW170817. We use the priors
shown in Table \ref{tab:parameters}, and take 4 s of data sampled at 2048 Hz
starting at 20 Hz for GW150914, and 128 s of data sampled at 4096 Hz starting
at 23 Hz for GW170817; strain data and PSDs for both events are fetched from
GWOSC \citep{GWOSC}. We use the same normalizing flow model as the
injection-recovery study. For our specific choice of sampler settings, we
produce ${\sim}$2500 and 3500 \emph{effective samples}\footnote{Effective
samples here refers to the number of independent samples, which is the total
number of generated samples divided by their correlation length; we compute the
effective sample size using \textsc{arviz}
\citep{arviz_2019}, \url{https://python.arviz.org/en/stable/api/generated/arviz.ess.html}.
} for GW150914 and GW170817 respectively. Running on an \textsc{Nvidia} A100
GPU, the wall time for both events is around 3 minutes. Most of this time is
spent on just-in-time (JIT) compilation of the code; the actual sampling time
is only ${\sim}40$ s. We pre-compute the reference waveform parameters used to
hetetrodyne the likelihood for the two events, which is omitted in the
wall-time calculation.

\begin{figure*}
    \includegraphics[width=0.99\textwidth]{figures/GW150914.pdf}
    \caption{
      GW150914 posterior computed by our code (blue) and \textsc{Bilby} (gray).
    }
    \label{fig:GW150914}
    \script{plot_GW150914.py}
\end{figure*}

\begin{figure*}
\includegraphics[width=0.99\textwidth]{figures/GW170817.pdf}
\caption{
    GW170817 posterior computed by our code (blue) and \textsc{Bilby} (gray).
}
\label{fig:GW170817}
    \script{plot_GW170817.py}
\end{figure*}

For comparison, we produce equivalent runs with \textsc{Bilby}, using the same
exact data and priors.  We use the \textsc{dynesty} sampler
\citep{2020MNRAS.493.3132S,dynesty}, with 1000 live points and other settings as
in the configurations files available on github \url{https://github.com/kazewong/TurboPE/tree/main/src}.  We
carry out these runs using \textsc{parallel Bilby} (\textsc{pBilby})
\citep{Smith:2019ucc} to distribute the computation over 400 Intel Skylake CPUs
for each event.  For the specific settings chosen, the wall-time duration of
each run was ${\sim}2$ h for GW150914 and ${\sim}1$ day for GW170817.

@@

Figs.~\ref{fig:GW150914} and \ref{fig:GW170817} show that our posteriors are
consistent with those produced by \textsc{Bilby}.  For a quantitative
comparison, we compute the Jensen-Shannon divergence between our code and
\textsc{Bilby} for the marginalized distribution for each parameter. The
Jensen-Shannon divergence (JSD) is a symmetric measure of the distance between
two probability distributions, with a value of 0 indicating identical
distributions and a value of $\ln{2}$ nat representing the maximum
possible divergence between two distributions. The JSD values for the two events
are shown in Table \ref{tab:JSD}. The maximum JSDs for GW150914 and GW170817 are
\jsdMaxBBH and \jsdMaxBNS, and the mean JSDs are \jsdAvgBBH and \jsdAvgBNS,
respectively. The JSD values are comparable to those reported in
\cite{Romero-Shaw:2020owr}, which show our code agrees with existing tools.

\begin{table}[hbt!]
    \begin{center}
    \variable{output/js_table.tex}
    \caption{JSD values for the marginalized distribution of each parameter for
    GW150914 and GW170817 between our code and \textsc{Bilby}. The bold values
    indicate the parameters with the largest JSD.}
    \label{tab:JSD}
    \end{center}
\end{table}

@@

% Additional traits for having differentiable waveforms
We are currently working on a number of improvements and extensions to our
current infrastructure. While the \textsc{IMRPhenomD} waveform approximant is a
reasonable start, it lacks some qualitative features that other
state-of-the-art models have, such as precession, subdominants moments of the
radiation, tides, or eccentricity. It also has a higher mismatch with reference
numerical relativity waveforms compared to more recent waveform models.
Currently, we are working on building differentiable implementations of
\textsc{IMRPhenomXPHM} \citep{Pratten:2020ceb}, a precessing successor to
\textsc{IMRPhenomD} including subdominant harmonics, as well as the numerical
relativity surrogate waveforms, including \textsc{NRSur7dq4}
\citep{Varma:2018mmi}. Going forward, we expect the use of autodifferentiation
environments like \textsc{Jax} to become more prevalent in the waveform
development community, increasing the number of differentiable waveform models
available. This would not only be beneficial for parameter estimation, but also
for a number of other applications such as Fisher matrix computations, template
placement and calibrating waveforms to numerical relativity \citep{Coogan:2022qxs,Iacovelli:2022bbs,Iacovelli:2022mbg,ripplepaper}.

@@

% JIT overhead
When it comes to wall time, the just-in-time compilation of our code is the
current limiting factor.  While \textsc{Jax}'s JIT compilation drastically
accelerates likelihood evaluations, it comes with a significant compilation
overhead before the first evaluation. We observe that the compilation time
depends on the device where the code is run; this is expected since
\textsc{Jax} leverages the \textsc{Accelerated Linear Algebra} (XLA) compiler
to take advantage of hardware accelerators, which means that \textsc{Jax} needs
to compile the code for each specific device according to its architecture. On
an \textsc{Nvidia} A100 GPU, the compilation overhead could go up to 3 minutes
for our current waveform. Meanwhile, for the cases we have studied, the time
needed to obtain converging sampling on an A100 is about $\sim 40\,$s. This
means the compilation overhead dominates the wall-clock time of our current PE
runs. To maximize the potential of our code, we are looking into ways to reduce
the compilation overhead or to cache the compilation results to avoid paying
the compilation overhead for every event.

% Faster maximum likelihood finder
Besides compilation, there is in principle also overhead from finding the
reference waveform used in for heterodyning the likelihood. Since the
\textsc{differential evolution} algorithm we currently use has not been
implemented in \textsc{Jax}, and the \textsc{Jax} waveform we use is not
compatible with the parallelization scheme in the \textsc{scipy} library,
maximizing the likelihood currently takes us around 1 minute for GW170817. There
are two ways to reduce this time.

@@

% Brief summary
In this work, we presented a PE pipeline for GW events that is efficient,
flexible and reliable. Our package brings together a number of innovations,
including differentiable waveform models, likelihood heterodyning, and
normalizing-flow enhanced gradient-based sampling. We tested the robustness of
our pipeline, currently built upon \textsc{ripple} and \textsc{flowMC}, on a
set of 1200 synthetic GW events, showing it is robust, unbiased and efficient
enough to handle the large catalogs of detections that will be available in the
near future. We also show that our pipeline can estimate the parameters of
GW150914 and GW170817 within a couple minutes, demonstrating the potential of our
implementation on real data.

>>>>

% Old codes have trouble handling next gen data
% The PE results in the most recent catalog of GW events are produced by several
% community-developed PE codes
% \mi{what catalog do you mean? The LVK doesn't use PyCBC Inference}
There exist a number of prominent, community-developed PE codes, including
\software{LALInference} \citep{Veitch:2014wba}, \software{PyCBC Inference}
\citep{Biwer:2018osg}, and \software{Bilby}
\citep{Ashton:2018jfp,Romero-Shaw:2020owr}.  These packages have been tested by a
number of groups and are well regarded as standard tools. However, while these
tools have passed many robustness tests, they are known to be computationally
intensive. The exact amount of time needed to analyze one event depends on
factors like the duration and frequency of the signal, as well as features of
the specific waveform model. Typical runtimes for production-level analyses can
range from hours to weeks.

@@

To obtain a reference waveform, we currently use the \software{differential
evolution} algorithm \citep{Storn1997DifferentialE} available in the
\software{scipy} package \citep{2020SciPy-NMeth} to find the waveform parameters
which maximize the likelihood.  The reference waveform could also be produced
from trigger parameters precomputed by a search pipeline without additional
computation. Once we have obtained a reference waveform, we can check the accuracy
of the heterodyned likelihood by comparing its value to the original likelihood
at several points in the parameter space. We can then choose the number of bins such
that the difference between the values of the two likelihoods is smaller than
chosen tolerance threshold.

@@

While transition kernels that use gradient information can help improve the
acceptance rate, computing the gradient of the posterior density function
introduces an additional computational cost, which is not necessarily
beneficial in terms of sampling time. If one wants to compute the gradient
through finite differencing, the additional computational cost goes as at least
${\sim} \mathcal{O}(2n)$, where $n$ is the dimension of the problem. On the
other hand, schemes like \software{Jax}
\citep{jax2018github} allow us to compute the gradient of the likelihood
function with respect to the parameters through automatic differentiation,
which gives the gradient information down to machine precision at around the
same order of time compared to evaluating the posterior itself. Thus, having
access to gradient information through automatic differentiation is crucial to
making gradient-based transition kernels favorable in terms of computing cost.

To leverage automatic differentiation, the entire posterior function must be
implemented within \software{Jax} or a similar framework; this includes the
likelihood and, therefore, the waveform approximant.  This means that the
development of differentiable approximants is essential for leveraging
gradient-based sampling in GW applications. We currently make use of the
waveforms implemented in \software{ripple} \citep{ripplepaper}.

@@

Note that every time we retrain the network, we are breaking the Markov
properties since we are changing the proposal distribution. To produce final
samples that can be used to estimate target quantities, one has to freeze the
normalizing flow model and not retrain during the final sampling phase in order
to satisfy the detailed balance condition. We use the package \software{flowMC}
\citep{2022arXiv221106397W,Gabrie:2021tlu}, with MALA as the gradient-based sampler. The
pseudocode of the algorithm is given in Algorithm \ref{alg:cap}.

\begin{algorithm}
\caption{\software{flowMC} pseudocode}\label{alg:cap}
\KwIn{initial position $ip$}
\Parameters{number of training loops $nt$, number of production loops $np$}
\Variables{current chains $cc$, current position $cp$, current NF parameters $\Theta$, chains from local sampler $c_{local}$, chains from global sampler $c_{global}$}
\KwResult{output chains $chains$}
$cp \leftarrow ip$\\
\tcc{Training loop}
    \For{$i<nt$}{
        $cc, cp \leftarrow LocalSampling(cp)$\\
        $\Theta \leftarrow TuneNF(cc)$\\
        $c_{global}, cp \leftarrow GlobalSampling(cp, \Theta)$ \\
        $cc \leftarrow Append(cc, c_{global})$
    }
\tcc{Production loop}
    \For{$i<np$}{
        $c_{local}, cp \leftarrow LocalSampling(cp)$\\
        $c_{global}, cp \leftarrow GlobalSampling(cp, \Theta)$ \\
        $chains \leftarrow Append(chains, c_{local}, c_{global})$
    }

@@

% GPU helps packing a shit ton of waveform evaluation
Another benefit of accelerators is the parallel evaluation of waveforms. Since
the waveform model we use can be evaluated at any given time or frequency
independently, this means computing a waveform can be trivially parallelized
over frequency bins. On an \textsc{Nvidia} A100 GPU, we can evaluate the
waveform model $\sim \mathcal{O}(10^9)$ times in a second for different
frequencies or source parameters. Together with the heterodyned likelihood, this
allows us to run thousands of parallel chains in a PE run. The high throughput
of waveform evaluations unlocks the potential of the \software{flowMC} sampling
algorithm.

@@

% State the distribution of injected population
We sample 1200 events from the distribution of parameters detailed in Table
\ref{tab:parameters}; the same distributions are used as the prior in the PE
process.  We simulate signals over $16\,$s of data, with a minimum frequency
cutoff of $30\,$Hz and a sampling rate of $2048\,$Hz. We draw noise from a set of
projected design PSDs for the LIGO Hanford, LIGO Livingston
(\texttt{SimNoisePSDaLIGOZeroDetHighPower}) and Virgo
(\texttt{SimNoisePSDAdvVirgo}) detectors
\citep{lalsuite,Shoemaker:T0900288,2012arXiv1202.4031M}. For both injection and
recovery, we make use of the \textsc{IMRPhenomD} waveform \citep{Khan:2015jqa}
via the fully-differentiable implementation presented in the \software{ripple}
package \citep{ripplepaper}. We use a neural spline flow model \citep{2019arXiv190604032D}
with 10 layers, each with 128 hidden units, and 8 bins per layer as our
normalizing flow model.

@@

To demonstrate the performance of our parameter estimation pipeline, we apply
it to two real LIGO-Virgo events: GW150914 and GW170817. We use the priors
shown in Table \ref{tab:parameters}, and take 4 s of data sampled at 2048 Hz
starting at 20 Hz for GW150914, and 128 s of data sampled at 4096 Hz starting
at 23 Hz for GW170817; strain data and PSDs for both events are fetched from
GWOSC \citep{GWOSC}. We use the same normalizing flow model as the
injection-recovery study. For our specific choice of sampler settings, we
produce ${\sim}$2500 and 3500 \emph{effective samples}\footnote{Effective
samples here refers to the number of independent samples, which is the total
number of generated samples divided by their correlation length; we compute the
effective sample size using \software{arviz}
\citep{arviz_2019}, \url{https://python.arviz.org/en/stable/api/generated/arviz.ess.html}.
} for GW150914 and GW170817 respectively. Running on an \textsc{Nvidia} A100
GPU, the wall time for both events is around 3 minutes. Most of this time is
spent on just-in-time (JIT) compilation of the code; the actual sampling time
is only ${\sim}40$ s. We pre-compute the reference waveform parameters used to
hetetrodyne the likelihood for the two events, which is omitted in the
wall-time calculation.

\begin{figure*}
    \includegraphics[width=0.99\textwidth]{figures/GW150914.pdf}
    \caption{
      GW150914 posterior computed by our code (blue) and \software{Bilby} (gray).
    }
    \label{fig:GW150914}
    \script{plot_GW150914.py}
\end{figure*}

\begin{figure*}
\includegraphics[width=0.99\textwidth]{figures/GW170817.pdf}
\caption{
    GW170817 posterior computed by our code (blue) and \software{Bilby} (gray).
}
\label{fig:GW170817}
    \script{plot_GW170817.py}
\end{figure*}

For comparison, we produce equivalent runs with \software{Bilby}, using the same
exact data and priors.  We use the \software{dynesty} sampler
\citep{2020MNRAS.493.3132S,dynesty}, with 1000 live points and other settings as
in the configurations files available on github \url{https://github.com/kazewong/TurboPE/tree/main/src}.  We
carry out these runs using \software{parallel Bilby} (\software{pBilby})
\citep{Smith:2019ucc} to distribute the computation over 400 Intel Skylake CPUs
for each event.  For the specific settings chosen, the wall-time duration of
each run was ${\sim}2$ h for GW150914 and ${\sim}1$ day for GW170817.

@@

Figs.~\ref{fig:GW150914} and \ref{fig:GW170817} show that our posteriors are
consistent with those produced by \software{Bilby}.  For a quantitative
comparison, we compute the Jensen-Shannon divergence between our code and
\software{Bilby} for the marginalized distribution for each parameter. The
Jensen-Shannon divergence (JSD) is a symmetric measure of the distance between
two probability distributions, with a value of 0 indicating identical
distributions and a value of $\ln{2}$ nat representing the maximum
possible divergence between two distributions. The JSD values for the two events
are shown in Table \ref{tab:JSD}. The maximum JSDs for GW150914 and GW170817 are
\jsdMaxBBH and \jsdMaxBNS, and the mean JSDs are \jsdAvgBBH and \jsdAvgBNS,
respectively. The JSD values are comparable to those reported in
\cite{Romero-Shaw:2020owr}, which show our code agrees with existing tools.

\begin{table}[hbt!]
    \begin{center}
    \variable{output/js_table.tex}
    \caption{JSD values for the marginalized distribution of each parameter for
    GW150914 and GW170817 between our code and \software{Bilby}. The bold values
    indicate the parameters with the largest JSD.}
    \label{tab:JSD}
    \end{center}
\end{table}

@@

% Additional traits for having differentiable waveforms
We are currently working on a number of improvements and extensions to our
current infrastructure. While the \textsc{IMRPhenomD} waveform approximant is a
reasonable start, it lacks some qualitative features that other
state-of-the-art models have, such as precession, subdominants moments of the
radiation, tides, or eccentricity. It also has a higher mismatch with reference
numerical relativity waveforms compared to more recent waveform models.
Currently, we are working on building differentiable implementations of
\textsc{IMRPhenomXPHM} \citep{Pratten:2020ceb}, a precessing successor to
\textsc{IMRPhenomD} including subdominant harmonics, as well as the numerical
relativity surrogate waveforms, including \textsc{NRSur7dq4}
\citep{Varma:2018mmi}. Going forward, we expect the use of autodifferentiation
environments like \software{Jax} to become more prevalent in the waveform
development community, increasing the number of differentiable waveform models
available. This would not only be beneficial for parameter estimation, but also
for a number of other applications such as Fisher matrix computations, template
placement and calibrating waveforms to numerical relativity \citep{Coogan:2022qxs,Iacovelli:2022bbs,Iacovelli:2022mbg,ripplepaper}.

@@

% JIT overhead
When it comes to wall time, the just-in-time compilation of our code is the
current limiting factor.  While \software{Jax}'s JIT compilation drastically
accelerates likelihood evaluations, it comes with a significant compilation
overhead before the first evaluation. We observe that the compilation time
depends on the device where the code is run; this is expected since
\software{Jax} leverages the \textsc{Accelerated Linear Algebra} (XLA) compiler
to take advantage of hardware accelerators, which means that \software{Jax} needs
to compile the code for each specific device according to its architecture. On
an \textsc{Nvidia} A100 GPU, the compilation overhead could go up to 3 minutes
for our current waveform. Meanwhile, for the cases we have studied, the time
needed to obtain converging sampling on an A100 is about $\sim 40\,$s. This
means the compilation overhead dominates the wall-clock time of our current PE
runs. To maximize the potential of our code, we are looking into ways to reduce
the compilation overhead or to cache the compilation results to avoid paying
the compilation overhead for every event.

% Faster maximum likelihood finder
Besides compilation, there is in principle also overhead from finding the
reference waveform used in for heterodyning the likelihood. Since the
\software{differential evolution} algorithm we currently use has not been
implemented in \software{Jax}, and the \software{Jax} waveform we use is not
compatible with the parallelization scheme in the \software{scipy} library,
maximizing the likelihood currently takes us around 1 minute for GW170817. There
are two ways to reduce this time.

@@

% Brief summary
In this work, we presented a PE pipeline for GW events that is efficient,
flexible and reliable. Our package brings together a number of innovations,
including differentiable waveform models, likelihood heterodyning, and
normalizing-flow enhanced gradient-based sampling. We tested the robustness of
our pipeline, currently built upon \software{ripple} and \software{flowMC}, on a
set of 1200 synthetic GW events, showing it is robust, unbiased and efficient
enough to handle the large catalogs of detections that will be available in the
near future. We also show that our pipeline can estimate the parameters of
GW150914 and GW170817 within a couple minutes, demonstrating the potential of our
implementation on real data.
